---
title: "Mortality in Denmark 2020"
author: "Thomas Poulsen"
date: "2021-01-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
#library(ggerr)
library(knitr)

CVpct <- function(x) 100*sd(x,na.rm=TRUE)/mean(x, na.rm=TRUE)
assert_rows <- function(n,x) {
  if(nrow(x) != n)
    stop(sprintf("Expected %s rows, got %s",n,nrow(x)))
  x
}

tb2.in <- read_excel("dødsfald_dk_2020-12-31_edit.xlsx", skip=2)
tb2 <- tidyr::pivot_longer(tb2.in, matches("\\d+U\\d+"), names_to = "week_id", values_to = "count") %>%
    mutate(year = as.numeric(sub("(\\d+)U(\\d+)$", "\\1", week_id)),
           week = as.numeric(sub("(\\d+)U(\\d+)$", "\\2", week_id)),
           age = ordered(Age, levels= unique(Age)),
           quarter = ceiling(week/13)
           )

tb2_mean <- tb2 %>% group_by(Region, Gender, Age, age , week) %>% summarise(week_mean = mean(count))

# population for normalization
#  https://www.statistikbanken.dk/statbank5a/selectvarval/define.asp?PLanguage=0&subword=tabsel&MainTable=FOLK1A&PXSId=199113&tablestyle=&ST=SD&buttons=0
pop.in <- read_excel("befolkning_2008-2020_edit.xlsx", skip=2)
pop1 <- tidyr::pivot_longer(pop.in[,-c(1,2)], matches("\\d+K\\d+"), names_to = "quarter_id", values_to="population") %>%
  mutate(year = as.numeric(sub("(\\d+)K(\\d+)$", "\\1", quarter_id)),
         quarter = as.numeric(sub("(\\d+)K(\\d+)$", "\\2", quarter_id))) 

tb3 <- merge(tb2, pop1, by=c("Gender","year","quarter")) %>%
  mutate(mortality_rate = count/population, mortality_per_mio = mortality_rate * 1E6)

tb3_mean <- tb3 %>% group_by(Region, Gender, Age, age , week) %>% summarise(week_mean = mean(mortality_per_mio))

```

# Conclusion

Despite the Covid-19 pandemic, total mortality in Denmark in 2020 is not different from the average over 2008-2020.
But this hides that mortality for people aged 40-70 was lower than normal, and mortality for ages above 70 was higher.

This analysis is inspired by a piece by Sonja Furu in [Weekendavisen 2020-53](https://www.weekendavisen.dk/2020-53/samfund/lys-i-moerket).

# Average mortality in Denmark is 181 per million per week.

The average mortality in Denmark in the period 2008-2020 is 181 (+/- 9%) per week per 1 million people.
It is about half a percent lower for females and half a percent higher for males.

Some legends are in Danish. The translation is:

* "I alt": Total
* "Mænd": Male
* "Kvinder": Female
* "år": years

```{r echo=FALSE}
tbl1 <- tb3 %>% subset(Age == "I alt") %>% group_by(Gender) %>% summarise(mortality_per_mio = mean(mortality_rate*1E6), CVpct = CVpct(mortality_rate), .groups="drop") %>% ungroup()

kable(tbl1)

kable(subset(tb3, Age == "I alt") %>% group_by(Gender) %>% summarise(population = mean(population), .groups="drop") %>% ungroup())

```


The plot below shows the weekly mortality over the span 2008 to 2020 as blue dots, and the 2020 value as a black line. 
This convention is used in most plots below.

There is no sign of a higher mortality in 2020.

```{r echo=FALSE}
ggplot(subset(tb3, Gender =="I alt" & Age == "I alt"), aes(x = week, y = mortality_per_mio)) + geom_point(color = "blue", alpha =.5) + geom_line(data = subset(tb3, year == 2020 & Gender =="I alt" & Age == "I alt"), aes(x = week, y = mortality_per_mio), color = "black") +  ggtitle("Mortality in Denmark per week 2008-2020")

```

# Most die at age 80-90

Mortality is highest for the age groups 80-85 and 85-89.

Average over 2008-2020 is a blue dot, 2020 value is a black line. 
There is no clear over mortality in 2020.


```{r echo=FALSE, fig.height=10, fig.width=10}
ggplot(subset(tb3_mean, Gender =="I alt" & Age != "I alt"), aes(x = week, y = week_mean)) + geom_point(color = "blue") + geom_line(data = subset(tb3, year == 2020 & Gender =="I alt" & Age != "I alt"), aes(x = week, y = mortality_per_mio), color = "black") + facet_wrap(~age)+ ggtitle("Mortality per mio in DK per week")

```

# In 2020 mortility was low for middle aged, and high for elderly.

The plot below splits the data age group. Note that each panel has a separate y-axes.

This shows some interesting effects.

Below 40 it looks most ĺy normal, except for 20-24 years that had da hefty summer.

From 40 to 70 years old mortality is lower than average, except for week 17 (age 45-49) and week 19 (age 50-54).

For ages from 70 to 99 there is a higher mortality in 2020, especially in the weeks between 10 and 20.
The lockdown in Denmark started week 11.


```{r echo=FALSE, fig.height=10, fig.width=10}
ggplot(subset(tb3_mean, Gender =="I alt" & Age != "I alt"), aes(x = week, y = week_mean)) + geom_point(color = "blue") + geom_line(data = subset(tb3, year == 2020 & Gender =="I alt" & Age != "I alt"), aes(x = week, y = mortality_per_mio), color = "black") + facet_wrap(~age, scale="free")+ ggtitle("Mortality per mio per week in DK")

```

## Gender effect

In general, male over-mortality starts early and is clear until age 80-85 years.
This is also the age group with highest male mortality. 
Women typically die at are 85-90.

Looking closer at the plots below, it looks like males are mainly driving the differnce of 2020 to the average.

In the plot below the line is 2020 value and the points are average 2008-2020. The color on both points and lines follow the gender.

```{r echo=FALSE, fig.height=10, fig.width=10}

ggplot(subset(tb3_mean, Gender !="I alt" & Age != "I alt"), aes(x = week, y = week_mean, color = Gender)) + geom_point() + geom_line(data = subset(tb3, year == 2020 & Gender !="I alt" & Age != "I alt"), aes(x = week, y = mortality_per_mio, color = Gender)) + facet_wrap(~age, scales = "free")+ ggtitle("Mortality per mio per week in DK")

ggplot(subset(tb3_mean, Gender =="Mænd" & Age != "I alt"), aes(x = week, y = week_mean)) + geom_point() + geom_line(data = subset(tb3, year == 2020 & Gender =="Mænd" & Age != "I alt"), aes(x = week, y = mortality_per_mio)) + facet_wrap(~age, scales = "free")+ ggtitle("Male Mortality per mio per week in DK")

ggplot(subset(tb3_mean, Gender == "Kvinder" & Age != "I alt"), aes(x = week, y = week_mean)) + geom_point() + geom_line(data = subset(tb3, year == 2020 & Gender =="Kvinder" & Age != "I alt"), aes(x = week, y = mortality_per_mio)) + facet_wrap(~age, scales = "free")+ ggtitle("Female Mortality per mio per week in DK")


```

# Data

Mortality numbers per week, age and gender 2007-2020 downloaded from 
Danmarks Statistik on 2020-12-31:  https://www.statistikbanken.dk/statbank5a/SelectVarVal/Define.asp?MainTable=DODC2&PLanguage=0&PXSId=0&wsid=cftree

Data is also available split on region and at a resolution of 1 day.

Population data for normalization per quarter from 2008 to 2020 was downloaded 2020-01-02 from  https://www.statistikbanken.dk/statbank5a/selectvarval/define.asp?PLanguage=0&subword=tabsel&MainTable=FOLK1A&PXSId=199113&tablestyle=&ST=SD&buttons=0
